Supplementary materials for: Modelling palaeoecological time series using generalized additive models

### 2018-08-16 supplementary-materials.tex
bump date
mass rebuild after switching to gratia package

  \title{Supplementary materials for: Modelling palaeoecological time series
using generalized additive models}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Gavin L. Simpson}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{August 14, 2018}

@@

In addition, the \emph{schoenberg} package is required; this
in-development package is not on CRAN but can be installed directly from
GitHub using functions from the \emph{devtools} package. To install the
package, install \emph{devtools} and then use
\texttt{devtools::install\_github()} to install \emph{schoenberg}, as
shown below:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## schoenberg is not on CRAN, install from github:}
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
\NormalTok{devtools::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"gavinsimpson/schoenberg"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Assuming that the installation of \emph{schoenberg} completes without
error, the package can be loaded as usual

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"schoenberg"}\NormalTok{) }\CommentTok{# need to change the name of the package}
\end{Highlighting}
\end{Shaded}

@@

\NormalTok{## Generate a plot of the data}
\NormalTok{braya_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{colour =} \StringTok{"grey"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}

@@

\begin{verbatim}
#>         lower      est.     upper
#> Phi 0.2811107 0.6026967 0.8547543
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}

@@

The same model is fitted to the Braya-Sø data set. Note however that in
order to even fit the model with both a smooth and the CAR(1) process, I
have had to change the default optimiser used to estimate the model
parameters, and reduce the basis dimension to a small number. We also
fit the model using GCV, which is the defaut, hence no \texttt{method}
argument

@@

\begin{verbatim}
#>            lower      est. upper
#> Phi 1.611049e-22 0.2000156     1
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}

@@

For Braya-Sø, we repeat the process, but we do so for both models (GAMM
+ CAR(1) and GCV), and use a critical value from the \(t\) distribution
to form the confidence interval

@@

Figure 5 in the manuscript was produced using:

@@

Clearly we are not able to identify both the wiggly trend and the CAR(1)
process in a single model. We might proceed by fitting a simple GAM via
\texttt{gam()} with REML smoothness selection and moderate number of
basis functions --- here we set \texttt{k\ =\ 15} for illustration.

@@

\begin{verbatim}
#> 
#> Method: REML   Optimizer: outer newton
#> full convergence after 6 iterations.
#> Gradient range [-4.468978e-08,8.696688e-10]
#> (score -187.2524 & scale 0.000689802).
#> Hessian positive definite, eigenvalue range [0.5525272,43.5153].
#> Model rank =  15 / 15 
#> 
#> Basis dimension (k) checking results. Low p-value (k-index<1) may
#> indicate that k is too low, especially if edf is close to k'.
#> 
#>            k'   edf k-index p-value    
#> s(Year) 14.00  2.62    0.56  <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

In the output, the column labelled \texttt{k\textquotesingle{}} is the
size of the basis used (the number of basis functions) --- note that it
is 14 and not the requested 15 functions as the constant basis function
has been removed as it is confounded with the model constant term, the
intercept. The column labelled \texttt{edf} indicates the
\emph{effective degrees of freedom} for the estimated smooth. The
\texttt{k-index} column contains the value of a test statistic for a
test of sufficient number of basis functions --- ideally, we'd like the
value of \texttt{k-index} to be close to or greater than 1. The quoted
\emph{p} value measures the support for the null hypothesis that enough
basis functions were available. In practial terms, the low
\texttt{k-index} and very low \emph{p} value suggest that the requested
number of basis functions used to fit the model was too low.

The solution is to increase \texttt{k}, say by doubling the original
value, and refit the model and perform the basis dimension check using
\texttt{gam.check()}, repeating this process until \texttt{k-index} is
close to or greater than 1 and the \emph{p} value is larger than a
desired threshold. In my experience, it may not be possible with some
data sets to increase \texttt{k} large enough to result in a
\texttt{k-index} \textgreater{} 1 and a \emph{p} value \textgreater{}
0.05, given the available. It is important to remember that the test
provided by \texttt{gam.check()} is only an heuristic one and shouldn't
be seen as infallible. If you find yourself increasing \texttt{k} to
large values relative to the number of samples in your data set without
achieving a \texttt{k-index} \textgreater{} 1 or \emph{p} \textgreater{}
0.05, then note the value in the \texttt{edf} column as you increase
\texttt{k}. If the \texttt{edf} of the model changes little as you
continue to increase \texttt{k}, this may be evidence that the test is
failing in this particular instance and that the estimated smooth is not
dependent on the value of \texttt{k} chosen.

@@

To proceed with the Braya-Sø example, we need to increase the basis
dimension (\texttt{k\ =\ 45}), fit using \texttt{method\ =\ "REML"}. As
discussed in the manuscript, we should also account for the non-constant
variance, or \emph{heteroscedasticity}, that may arise due to variation
in the amount of time (or ``lake years'') that each sample represents.
All else equal, we expect that samples that average more time have
smaller variance than those that average a smaller amount of time. To
include information about the expected heteroscedasticity in the GAM we
can use observational weights\footnote{This simple solution is for a
  Gaussian GAM. For models with other distributions, the exact
  interpretation of observational weights will vary. For such data,
  where the conditional mean and variance of the observations are
  linked, the use of location-scale, or distributional, models may be
  more apporpriate and informative.}. Here I use the
\texttt{sampleInterval} variable that I created earlier as the measure
of lake years per sample, and to avoid changing the model likelihood,
the weights use are the values of \texttt{sampleInterval} divided by the
mean of \texttt{sampleInterval}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
                  \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
\KeywordTok{summary}\NormalTok{(braya_reml)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> 
#> Family: gaussian 
#> Link function: identity 
#> 
#> Formula:
#> UK37 ~ s(Year, k = 45)
#> 
#> Parametric coefficients:
#>              Estimate Std. Error t value Pr(>|t|)    
#> (Intercept) -0.633734   0.001901  -333.3   <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Approximate significance of smooth terms:
#>           edf Ref.df     F p-value    
#> s(Year) 28.49  33.99 7.755  <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> R-sq.(adj) =  0.751   Deviance explained = 83.2%
#> -REML = -175.68  Scale est. = 0.00024036  n = 89
\end{verbatim}

If we now check if 44 (\texttt{k\ =\ 45}) basis functions is sufficient

@@

\begin{verbatim}
#> 
#> Method: REML   Optimizer: outer newton
#> full convergence after 7 iterations.
#> Gradient range [-2.210576e-06,-5.181457e-08]
#> (score -175.684 & scale 0.0002403604).
#> Hessian positive definite, eigenvalue range [2.524305,48.11146].
#> Model rank =  45 / 45 
#> 
#> Basis dimension (k) checking results. Low p-value (k-index<1) may
#> indicate that k is too low, especially if edf is close to k'.
#> 
#>           k'  edf k-index p-value
#> s(Year) 44.0 28.5    1.27    0.99
\end{verbatim}

we note that the value of \texttt{k-index} is now greater than 1 and the
\emph{p} value suggests there is very little evidence against the null
hypothesis. Also note that the \texttt{edf} of the estimated smooth is
considerably below the maximum (\texttt{k\textquotesingle{}}). This is
the effect of the smoothness penalty removing the excessive wiggliness
possible with 44 basis functions. If the \texttt{edf} were close to
\texttt{k\textquotesingle{}}, you might consider increasing \texttt{k}
by a modest amount (in this instance by 5 or 10 additional basis
functions) to assure yourself that you have sufficient basis functions.
This has the additional advantage of allowing a richer set of smooths of
a given complexity (\texttt{edf}) to be represented using the basis
functions, which may improve the model fit to the data without changing
the \texttt{edf} too much.

@@

Samples from the posterior distribution of a GAM can be drawn using the
\texttt{simulate()} methods from the \emph{schoenberg} package.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## posterior simulation}
\NormalTok{## need to reset-up newBraya}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya,}
                 \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                       \DataTypeTok{length.out =} \NormalTok{N)))}
\NormalTok{braya_pred <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
                    \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(braya_reml, newBraya,}
                                       \DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}

\NormalTok{## simulate}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(braya_reml, }\DataTypeTok{nsim =} \NormalTok{nsim, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
                  \DataTypeTok{unconditional =} \OtherTok{TRUE}\NormalTok{)}
\KeywordTok{colnames}\NormalTok{(sims2) <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"sim"}\NormalTok{, }\KeywordTok{seq_len}\NormalTok{(nsim))}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{stack}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(sims2)),}
                      \KeywordTok{c}\NormalTok{(}\StringTok{"simulated"}\NormalTok{, }\StringTok{"run"}\NormalTok{))}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(sims2, }\DataTypeTok{Year =} \KeywordTok{rep}\NormalTok{(newBraya$Year, nsim),}
                       \DataTypeTok{simulated =} \NormalTok{simulated)}

@@

Figure 7 in the manuscript was prepared using

@@

Across-the-function and simultaneous confidence intervals are computed
using the \texttt{confint()} method. The type of interval required is
given via the \texttt{type} argument with options \texttt{"confidence"}
and \texttt{"simultaneous"}. Fro example, for Small Water we would use

@@

\begin{verbatim}
#>    smooth     Year      est        se     crit    lower    upper
#> 1 s(Year) 1846.319 3.787215 0.1616144 3.087688 3.288200 4.286230
#> 2 s(Year) 1847.128 3.789747 0.1509100 3.087688 3.323784 4.255710
#> 3 s(Year) 1847.937 3.792304 0.1411439 3.087688 3.356496 4.228112
#> 4 s(Year) 1848.745 3.794914 0.1324961 3.087688 3.385807 4.204020
#> 5 s(Year) 1849.554 3.797602 0.1251070 3.087688 3.411310 4.183893
#> 6 s(Year) 1850.363 3.800394 0.1190552 3.087688 3.432788 4.167999
\end{verbatim}

@@

The first derivative of the estimated trend is calculated using finite
differences using the \texttt{fderiv()} function. There is also a
\texttt{confint()} method for objects produced by \texttt{fderiv()}. The
first derivatives and a 95\% simultaneous confidence interval for the
Small Water trend we computed and plotted using

@@

@@

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure 9-1} \end{center}

@@

For the Gaussian process smooth to fit within the GAM framework
described in the manuscript, we need to supply the value of \(\phi\) for
the effective range of the correlation function. To estimate \(\phi\),
we need to repeatedly fit the required GAM using a range of plausible
values for \(\phi\), which we do using a loop. In the chunk below I fit
200 models with \(\phi\) in the range 15--500. For value of \(\phi\) I
fit the required GAM and extract the REML score (from component
\texttt{gcv.ubre}) and store it in the numeric vectors \texttt{Mat} or
\texttt{SEx}, for Matérn and Squared Exponential correlation functions,
respectively. The final line of the chunk prepares the REML scores for
the two correlation functions in long or tidy format suitable for
plotting with \emph{ggplot2}.

@@

whilst Figure 12 in the manuscript was prepared using

@@

We write a small function to predict from each model over the range of
\texttt{Year} and return the data in tidy format for plotting. The plot
produce reproduces figure 13 in the manuscript.

@@

\NormalTok{pltData <-}\StringTok{ }\KeywordTok{do.call}\NormalTok{(}\StringTok{"rbind"}\NormalTok{, }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{list}\NormalTok{(plts_gp, plts_ad, plts_tprs),}
                                   \StringTok{`}\DataTypeTok{[[}\StringTok{`}\NormalTok{, }\StringTok{"objects"}\NormalTok{))}
\NormalTok{pltData <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pltData, }\DataTypeTok{Model =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GP"}\NormalTok{, }\StringTok{"Adaptive"}\NormalTok{, }\StringTok{"TPRS"}\NormalTok{),}
                              \DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(plts_gp$objects)))}

@@

@@

To make this section self-contained, I refitted the Small Water GAM plus
CAR(1) model

@@

and then we load the \textsuperscript{210}Pb dating results for the
dated core sections.

@@

The next chunk does the final step in the process. For each of the
models we just fitted to include age model uncertainty, we simulate 50
draws from the model posterior distribution. We start with a wrapper
function around the \texttt{simulate()} code we want to run on each
model, then do the actual posterior draws for each model using
\texttt{lapply()}. The final step just arranges data for plotting.

@@

\begin{verbatim}
#>  setting  value                                      
#>  version  R version 3.4.4 Patched (2018-03-17 r74422)
#>  system   x86_64, linux-gnu                          
#>  ui       X11                                        
#>  language (EN)                                       
#>  collate  en_CA.UTF-8                                
#>  tz       America/Regina                             
#>  date     2018-05-14
\end{verbatim}

>>>>

  \title{Supplementary materials for: Modelling palaeoecological time series
using generalized additive models}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Gavin L. Simpson}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{August 16, 2018}

@@

In addition, the \emph{gratia} package is required; this in-development
package is not on CRAN but can be installed directly from GitHub using
functions from the \emph{devtools} package. To install the package,
install \emph{devtools} and then use
\texttt{devtools::install\_github()} to install \emph{gratia}, as shown
below:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## gratia is not on CRAN, install from github:}
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
\NormalTok{devtools::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"gavinsimpson/gratia"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Assuming that the installation of \emph{gratia} completes without error,
the package can be loaded as usual

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"gratia"}\NormalTok{) }\CommentTok{# need to change the name of the package}
\end{Highlighting}
\end{Shaded}

@@

\NormalTok{## plot Braya-So data}
\NormalTok{braya_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{colour =} \StringTok{"grey"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}

@@

\begin{verbatim}
#>        lower      est.     upper
#> Phi 0.282388 0.6026967 0.8539689
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}

@@

The same model as used for Small Water was also initially attempted with
the Braya-Sø data. However, in order to even fit the model with both a
smooth and the CAR(1) process, I had to change the default optimizer
used to estimate the model parameters, and reduce the basis dimension,
\texttt{k}, to a small number. I also fit the model using GCV, which is,
at the time of writing, the default in \texttt{gam()}, hence no
\texttt{method} argument

@@

\begin{verbatim}
#>            lower      est. upper
#> Phi 2.680033e-31 0.2000156     1
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}

@@

To plot the GAMs fitted to the Braya-Sø time series, we repeat the
process used with the Small Water GAM, but we do this for both models
(GAMM + CAR(1) and GCV), using a critical value from the \(t\)
distribution to form the confidence interval

@@

Figure 6 in the manuscript was produced using:

@@

For the Braya-Sø data, clearly we are not able to identify both the
wiggly trend and the CAR(1) process in a single model. We proceed by
fitting a simple GAM via \texttt{gam()} with REML smoothness selection
and a moderate number of basis functions --- here we set
\texttt{k\ =\ 15} for illustration.

@@

\begin{verbatim}
#> 
#> Method: REML   Optimizer: outer newton
#> full convergence after 6 iterations.
#> Gradient range [-4.468978e-08,8.696768e-10]
#> (score -187.2524 & scale 0.000689802).
#> Hessian positive definite, eigenvalue range [0.5525272,43.5153].
#> Model rank =  15 / 15 
#> 
#> Basis dimension (k) checking results. Low p-value (k-index<1) may
#> indicate that k is too low, especially if edf is close to k'.
#> 
#>            k'   edf k-index p-value    
#> s(Year) 14.00  2.62    0.56  <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

The first few lines of output from \texttt{gam.check()} include
information on the model fit, whether the algorithm converged, and some
diagnostics from the optimization at convergence --- you are unlikely to
need this information, but it can be useful in diagnosing problems with
model fitting or when reporting problems to Simon Wood, the developer of
the \emph{mgcv} package. In the table at the bottom of the output from
\texttt{gam.check()}, the column labelled \texttt{k\textquotesingle{}}
is the size of the basis used to fit the model (the number of basis
functions). Note that the stated value is 14 and not the requested 15
functions because the constant basis function has been removed due to it
being confounded with the model constant term, the intercept. The column
labelled \texttt{edf} indicates the \emph{effective degrees of freedom}
for the estimated smooth, which is a measure of the complexity or
wiggliness of the final smooth. The \texttt{k-index} column contains the
value of a test statistic for a test of a sufficient number of basis
functions --- ideally, we'd like the value of \texttt{k-index} to be
close to or greater than 1. The quoted \emph{p} value measures the
support for the null hypothesis that enough basis functions were
available. In practial terms, the low \texttt{k-index} and very low
\emph{p} value shown above strongly suggest that the requested number of
basis functions used to fit the model was too low.

The solution is to increase \texttt{k}, say by doubling the original
value, and refit the model and perform the basis dimension check using
\texttt{gam.check()}, repeating this process until \texttt{k-index} is
close to or greater than 1 and the \emph{p} value is larger than a
desired threshold. In my experience fitting GAMs to palaeoenvironmental
time series, it may not be possible with some data sets to increase
\texttt{k} large enough to result in a \texttt{k-index} \textgreater{} 1
and a \emph{p} value \textgreater{} 0.05, given the available number of
samples. It is important to remember that the test provided by
\texttt{gam.check()} is only an heuristic one and shouldn't be seen as
infallible. If you find yourself increasing \texttt{k} to large values
relative to the number of samples in your data set without achieving a
\texttt{k-index} \textgreater{} 1 or \emph{p} \textgreater{} 0.05, then
note the value in the \texttt{edf} column as you increase \texttt{k}. If
the \texttt{edf} of the model changes little as you continue to increase
\texttt{k}, this may be evidence that the test is failing in this
particular instance and that the estimated smooth is not dependent on
the value of \texttt{k} chosen.

@@

To proceed with the Braya-Sø example, we need to increase the basis
dimension, fit using \texttt{method\ =\ "REML"}. As discussed in the
manuscript, we should also account for the non-constant variance, or
\emph{heteroscedasticity}, that may arise due to variation in the amount
of time (or ``lake years'') that each sample represents. All else equal,
we expect that samples that average more time have lower variance than
those that average a smaller amount of time. To include information
about the expected heteroscedasticity in the GAM we can use
observational weights\footnote{This simple solution is for a Gaussian
  GAM. For models with other distributions, the exact interpretation of
  observational weights will vary. For such data, where the conditional
  mean and variance of the observations are linked, the use of
  location-scale, or distributional, models may be more apporpriate and
  informative.}. Here, I use the \texttt{sampleInterval} variable that I
created earlier as the measure of lake years per sample, and, to avoid
changing the model likelihood, the weights use are the values of
\texttt{sampleInterval} divided by the mean of \texttt{sampleInterval}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{40}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
                  \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
\KeywordTok{summary}\NormalTok{(braya_reml)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> 
#> Family: gaussian 
#> Link function: identity 
#> 
#> Formula:
#> UK37 ~ s(Year, k = 40)
#> 
#> Parametric coefficients:
#>              Estimate Std. Error t value Pr(>|t|)    
#> (Intercept) -0.633741   0.001929  -328.5   <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Approximate significance of smooth terms:
#>           edf Ref.df     F p-value    
#> s(Year) 27.45  32.36 7.782  <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> R-sq.(adj) =  0.744   Deviance explained = 82.4%
#> -REML = -175.61  Scale est. = 0.00024757  n = 89
\end{verbatim}

If we now check if 39 (\texttt{k\ =\ 40}) basis functions is sufficient

@@

\begin{verbatim}
#> 
#> Method: REML   Optimizer: outer newton
#> full convergence after 9 iterations.
#> Gradient range [-1.105422e-06,1.500959e-07]
#> (score -175.6093 & scale 0.0002475669).
#> Hessian positive definite, eigenvalue range [2.644434,47.78257].
#> Model rank =  40 / 40 
#> 
#> Basis dimension (k) checking results. Low p-value (k-index<1) may
#> indicate that k is too low, especially if edf is close to k'.
#> 
#>           k'  edf k-index p-value
#> s(Year) 39.0 27.5    1.27    0.99
\end{verbatim}

we note that the value of \texttt{k-index} is now greater than 1 and the
\emph{p} value suggests there is very little evidence against the null
hypothesis. This result strongly suggests that there were sufficient
basis functions used to estimated the trend.

Also note that the \texttt{edf} of the estimated smooth is considerably
below the maximum (\texttt{k\textquotesingle{}}). This is the effect of
the smoothness penalty removing the excessive wiggliness possible with
39 basis functions. If the \texttt{edf} were close to
\texttt{k\textquotesingle{}}, you might consider increasing \texttt{k}
by a modest amount (in this instance by 5 or 10 additional basis
functions) to assure yourself that you have sufficient basis functions.
This has the additional advantage of allowing a richer set of smooths of
a given complexity (\texttt{edf}) to be represented using this larger
set of basis functions, which may improve the model fit to the data
without changing the \texttt{edf} too much.

@@

Samples from the posterior distribution of a GAM can be drawn using the
\texttt{simulate()} methods from the \emph{gratia} package.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## data points to simulate at}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya,}
                 \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                       \DataTypeTok{length.out =} \NormalTok{N)))}
\NormalTok{braya_pred <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
                    \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(braya_reml, newBraya,}
                                       \DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}

\NormalTok{## simulate}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(braya_reml, }\DataTypeTok{nsim =} \NormalTok{nsim, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
                  \DataTypeTok{unconditional =} \OtherTok{TRUE}\NormalTok{)}

\NormalTok{## rearrange the output into a long/tidy format}
\KeywordTok{colnames}\NormalTok{(sims2) <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"sim"}\NormalTok{, }\KeywordTok{seq_len}\NormalTok{(nsim))}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{stack}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(sims2)),}
                  \KeywordTok{c}\NormalTok{(}\StringTok{"simulated"}\NormalTok{, }\StringTok{"run"}\NormalTok{))}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(sims2, }\DataTypeTok{Year =} \KeywordTok{rep}\NormalTok{(newBraya$Year, nsim),}
                   \DataTypeTok{simulated =} \NormalTok{simulated)}

@@

@@

Across-the-function and simultaneous confidence intervals are computed
using the \texttt{confint()} method. The type of interval required is
given via the \texttt{type} argument with options \texttt{"confidence"}
and \texttt{"simultaneous"}. For example, for Small Water we would use

@@

\begin{verbatim}
#>    smooth     Year      est        se     crit    lower    upper
#> 1 s(Year) 1846.319 3.787215 0.1616144 3.072403 3.290671 4.283760
#> 2 s(Year) 1847.128 3.789747 0.1509100 3.072403 3.326090 4.253403
#> 3 s(Year) 1847.937 3.792304 0.1411439 3.072403 3.358653 4.225955
#> 4 s(Year) 1848.745 3.794914 0.1324961 3.072403 3.387832 4.201995
#> 5 s(Year) 1849.554 3.797602 0.1251070 3.072403 3.413222 4.181981
#> 6 s(Year) 1850.363 3.800394 0.1190552 3.072403 3.434608 4.166180
\end{verbatim}

@@

The first derivative of the estimated trend is calculated using finite
differences using the \texttt{fderiv()} function. There is also a
\texttt{confint()} method for objects produced by \texttt{fderiv()}. The
first derivatives and a 95\% simultaneous confidence interval for the
Small Water trend were computed and plotted using

@@

Figure 10 in the manuscript was prepared using

@@

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure-10-1} \end{center}

@@

For the Gaussian process smooth to fit within the GAM framework
described in the manuscript, we need to supply the value of \(\phi\) for
the effective range of the correlation function. To estimate \(\phi\),
we need to repeatedly fit the required GAM using a range of plausible
values for \(\phi\), which we do using a loop. In the chunk below I fit
200 models with \(\phi\) in the range 15--500. For each value of
\(\phi\) I fit the required GAM and extract the REML score (from
component \texttt{gcv.ubre}) and store it in the numeric vectors
\texttt{Mat} or \texttt{SEx}, for Matérn and Squared Exponential
correlation functions, respectively. The final line of the chunk
prepares the REML scores in long or tidy format suitable for plotting
with \emph{ggplot2}.

@@

whilst Figure 13 in the manuscript was prepared using

@@

I wrote a small function to predict from each model over the range of
\texttt{Year} and return the data in tidy format for plotting.

@@

\NormalTok{pltData <-}\StringTok{ }\KeywordTok{do.call}\NormalTok{(}\StringTok{"rbind"}\NormalTok{, }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{list}\NormalTok{(plts_gp, plts_ad, plts_tprs),}
                                   \StringTok{`}\DataTypeTok{[[}\StringTok{`}\NormalTok{, }\StringTok{"objects"}\NormalTok{))}
\NormalTok{pltData <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pltData,}
                     \DataTypeTok{Model =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GP"}\NormalTok{, }\StringTok{"Adaptive"}\NormalTok{, }\StringTok{"TPRS"}\NormalTok{),}
                                 \DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(plts_gp$objects)))}

@@

The plot produced reproduces Figure 14 in the manuscript.

@@

To make this section self-contained, I refitted the Small Water GAM plus
CAR(1) model. Because we will be sampling ages for each observation from
the posterior of the age-model GAM, the oldest sample will will be
somewhat younger than the expected value from the model in some samples.
This will cause problems when we come to simulate from the GAMs fitted
to the resampled age-models as each GAM will cover a slightly different
range of time. To avoid this problem I fix the knots at the extremes of
the observed values of \texttt{Year} and spread the remaining 12 knots
evenly inbetween. The knot locations are stored in the vector
\texttt{knots}, which is passed to the argument \texttt{knots} when
fitting the GAM

@@

Setting the knots like this doesn't change the model estimated above;
I'm only repeating what happens internally if you don't supply
\texttt{knots}. However, later we will need to specify this set of knots
explicitly when accounting for age model uncertainty.

Next we load the \textsuperscript{210}Pb dating results for the dated
core sections.

@@

The next chunk does the final step in the process. For each of the
models we just fitted, we simulate 50 draws from the model posterior
distribution. We start with a wrapper function around the
\texttt{simulate()} code we want to run on each model, then do the
actual posterior draws for each model using \texttt{lapply()}. The final
step just arranges data for plotting.

@@

\begin{verbatim}
#>  setting  value                                      
#>  version  R version 3.4.3 Patched (2018-01-16 r74126)
#>  system   x86_64, linux-gnu                          
#>  ui       X11                                        
#>  language (EN)                                       
#>  collate  en_CA.UTF-8                                
#>  tz       America/Regina                             
#>  date     2018-08-16
\end{verbatim}
